On existence of integral point sets and their diameter bounds \footnote{ This work was carried out at Voronezh State University and supported by the Russian Science Foundation grant 19-11-00197. }

### 2020-01-02 pps-linear-bound/pps-linear-lower-bound.tex
[J220] Fix minor remarks 1 to 7 from Review 2

For every finite point set, its diameter is naturally defined as
\begin{equation}
	\operatorname{diam} M = \max_{A,B\in M} |AB|
	.
\end{equation}
Another emerging question is: how does the diameter of an integral point set depend on its cardinality?
One can easily see that every $M\in\mathfrak{M}(m,n)$ with $\operatorname{diam} M = h$
can be dilated to $M_p\in\mathfrak{M}(m,n)$ with $\operatorname{diam} M = ph$
for every $p\in\mathbb{N}$.
So, the above question should be rephrased:
how does \textit{the least possible} diameter of an integral point set depend on its cardinality?
In order to answer this question, the following function was introduced~\cite{kurz2008bounds,kurz2008minimum}:
\begin{equation}
	d(m,n) = \min_{M\in\mathfrak{M}(m,n)} \operatorname{diam} M = \min_{M\in\mathfrak{M}(m,n)} \max_{A,B\in M} |AB|
	.
\end{equation}
We also refer to~\cite{kurz2008bounds} for a list of known exact values of $d(m,n)$
and its bounds; in the present paper, the case of $m=2$ will mostly be in the focus.

@@

Assuming that the planar integral point set contains many collinear points,
the following result holds.
\begin{theorem}~\cite[Theorem 4]{kurz2008minimum}
	For $\delta > 0$, $\varepsilon > 0$, and $P\in\mathfrak{M}(2,n)$ with
	at least $n^\delta$ collinear points there exists a $n_0 (\varepsilon)$
	such that for all $n \geq n_0 (\varepsilon)$ we have
	\begin{equation}
		\operatorname{diam} P \geq n^{\frac{\delta}{4 \log 2(1+\varepsilon)}\log \log n}
		.
	\end{equation}
\end{theorem}
However, the estimate is rather unsuitable for our needs,
as it does not provide the values of all the constants.
To obtain the needed estimate, we now prove a generalization of~\cite[lemma 3]{our-vmmsh-2018}.

@@

\begin{lemma}
	\label{lem:2k-1_segments}
	Let $M \in \mathfrak{M}(2,\mathbb{N})$ and let $m$ be a straight line.
	Then for every $k\in\mathbb{N}$ there are at most $2k-1$ segments $\Delta_i \subset m$ with endpoints in $M$,
	such that $|\Delta_i| = k$.
\end{lemma}
\begin{proof}
	Consider a point $N\in M \setminus m$.
	Then for each $\Delta_i$ there is a $\rho(n_i,\Delta_i)$-curve containing $N$.
	Due to Proposition~\ref{obs:rho_curves}, all $n_i$ are distinct;
	otherwise the $\rho(n_i,\Delta_i)$-curve and the $\rho(n_i,\Delta_j)$-curve, $j\neq i$, do not intersect.
	There can be only $2k-1$ distinct values for $n_i$,
	so there are at most $2k-1$ distinct segments $\Delta_i$.
\end{proof}

Setting $k=1$ in Lemma~\ref{lem:2k-1_segments},
we obtain the following result (which is exactly~\cite[Lemma 3]{our-vmmsh-2018}).
\begin{corollary}
	\label{cor:only_one_distance_1_on_straight_line}
	Let $M \in \mathfrak{M}(2,\mathbb{N})$ and let $m$ be a straight line.
	Then there is at most one pair of points $M_1,M_2\in M \cap m$
	such that $|M_1 M_2| = 1$.
\end{corollary}

@@

By means of some simple manipulations with the upper bound in Theorem~\ref{thm:varphi_k_bounds},
one can derive the following proposition.
\begin{proposition}
	\label{varphi_n_where_n_geq_21492}
	For $n \geq 21492$ we have
	\begin{equation}
		\varphi_n \leq \varphi_{n-1} \leq \frac{\beta}{\sqrt{n-2}}
		,
	\end{equation}
	where
	\begin{equation}
		\beta = \frac{1}{\sqrt{21490}} + \sqrt{ \frac{2}{\sqrt{3}} + \frac{1}{21490} } < 1.07464
		.
	\end{equation}
\end{proposition}

@@

In this section,
we improve the lower bound for minimum diameter of planar integral point sets
employing the results of two previous sections.

@@

We can improve the constant in Theorem~\ref{thm:main_estimate},
if some new exact values of $d(2,n)$ are found;
however, the obtained constant is bounded by $3^{1/4}\cdot2^{-3/2}$ above.
More precisely, we have the following theorem.

@@

Using the ``blowing up'' procedure described in~\cite[theorem 1.3]{kurz2008bounds},
we can construct an integral point set $M$ in $m$-dimensional Euclidean space, $m\geq 3$,
with distance 1 occuring in it.
If we have an integral point set with distance 1 occuring,
then we can easily dilate it to turn 1 into the desired distance.
These facts, together with Theorem~\ref{thm:minimeter_1_planar}, give the following theorem.

>>>>

For every finite point set, its diameter is naturally defined as
\begin{equation}
	\operatorname{diam} M = \max_{A,B\in M} |AB|
	.
\end{equation}
Another emerging question is: how does the diameter of an integral point set depend on its cardinality?
One can easily see that every $M\in\mathfrak{M}(m,n)$ with $\operatorname{diam} M = h$
can be dilated to $M_p\in\mathfrak{M}(m,n)$ with $\operatorname{diam} M = ph$
for every $p\in\mathbb{N}$.
So, the above question should be rephrased:
how does \textit{the least possible} diameter of an integral point set depend on its cardinality?
In order to answer this question, the following function was introduced~\cite{kurz2008bounds,kurz2008minimum}:
\begin{equation}
	d(m,n) = \min_{M\in\mathfrak{M}(m,n)} \operatorname{diam} M = \min_{M\in\mathfrak{M}(m,n)} \max_{A,B\in M} |AB|
	.
\end{equation}
We also refer to~\cite{kurz2008bounds} for a list of known exact values of $d(m,n)$ and its bounds.
In the present paper, the case of $m=2$ will mostly be in the focus.

@@

Assuming that the planar integral point set contains many collinear points,
the following result holds.
\begin{theorem}~\cite[Theorem 4]{kurz2008minimum}
	For $\delta > 0$, $\varepsilon > 0$, and $P\in\mathfrak{M}(2,n)$ with
	at least $n^\delta$ collinear points there exists a $n_0 (\varepsilon)$
	such that for all $n \geq n_0 (\varepsilon)$ we have
	\begin{equation}
		\operatorname{diam} P \geq n^{\frac{\delta}{4 \log 2(1+\varepsilon)}\log \log n}
		.
	\end{equation}
\end{theorem}
However, the estimate is rather unsuitable for our needs,
as it does not provide the values of all the constants.
To obtain the needed estimate, we now prove a generalization of~\cite[Lemma 3]{our-vmmsh-2018}.

@@

\begin{lemma}
	\label{lem:2k-1_segments}
	Let $M \in \mathfrak{M}(2,\mathbb{N})$ and let $l$ be a straight line.
	Then for every $k\in\mathbb{N}$ there are at most $2k-1$ segments $\Delta_i \subset l$ with endpoints in $M$,
	such that $|\Delta_i| = k$.
\end{lemma}
\begin{proof}
	Consider a point $N\in M \setminus l$.
	Then for each $\Delta_i$ there is a $\rho(n_i,\Delta_i)$-curve containing $N$.
	Due to Proposition~\ref{obs:rho_curves}, all $n_i$ are distinct;
	otherwise the $\rho(n_i,\Delta_i)$-curve and the $\rho(n_i,\Delta_j)$-curve, $j\neq i$, do not intersect.
	There can be only $2k-1$ distinct values for $n_i$,
	so there are at most $2k-1$ distinct segments $\Delta_i$.
\end{proof}

Setting $k=1$ in Lemma~\ref{lem:2k-1_segments},
we obtain the following result (which is exactly~\cite[Lemma 3]{our-vmmsh-2018}).
\begin{corollary}
	\label{cor:only_one_distance_1_on_straight_line}
	Let $M \in \mathfrak{M}(2,\mathbb{N})$ and let $l$ be a straight line.
	Then there is at most one pair of points $M_1,M_2\in M \cap l$
	such that $|M_1 M_2| = 1$.
\end{corollary}

@@

Now we need to estimate $\varphi_n$ with $\beta/\sqrt{n-n_0}$.
Using the upper bound of Theorem~\ref{thm:varphi_k_bounds},
one can easily find the values of $\beta$ and $n_0$.

\begin{proposition}
	\label{varphi_n_where_n_geq_21492}
	For $n \geq 21492$ we have
	\begin{equation}
		\varphi_n \leq \varphi_{n-1} \leq \frac{\beta}{\sqrt{n-2}}
		,
	\end{equation}
	where
	\begin{equation}
		\beta = \frac{1}{\sqrt{21490}} + \sqrt{ \frac{2}{\sqrt{3}} + \frac{1}{21490} } < 1.07464
		.
	\end{equation}
\end{proposition}

@@

In this section,
we improve the lower bound for the minimum diameter of planar integral point sets
employing the results of two previous sections.

@@

We can improve the constant in Theorem~\ref{thm:main_estimate},
if some new exact values of $d(2,n)$ are found;
however, the obtained constant is bounded by $3^{1/4}\cdot2^{-3/2}$ from above.
More precisely, we have the following theorem.

@@

Using the ``blowing up'' procedure described in~\cite[Theorem 1.3]{kurz2008bounds},
we can construct an integral point set $M$ in $m$-dimensional Euclidean space, $m\geq 3$,
with distance 1 occuring in it.
If we have an integral point set with distance 1 occuring,
then we can easily dilate it to turn 1 into the desired distance.
These facts, together with Theorem~\ref{thm:minimeter_1_planar}, give the following theorem.
